---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, Download_Load_Data}
library(plyr)

## Download the raw data
datadir <- "./datadir/"
zipfilename <- "./datadir/repdata_data_activity.zip"
filename <- "./datadir/activity.csv"
if (!file.exists(zipfilename)){
        dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(dataUrl, zipfilename, method="curl")
}  
## Extract data files.
        unzip(zipfile=zipfilename, exdir = datadir)
## Read data from csv file.
        raw_data <- read.csv(filename, stringsAsFactors = TRUE)
## Clean data, remove NA values from dataset.
        clean_data <- raw_data[complete.cases(raw_data),]
## Reorder colums to satisfy my personal taste.
        clean_data <- clean_data[c(2, 3, 1)]
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day.

```{r}
## Compute the total number of steps per day.
        total_steps_per_day_df <- ddply(clean_data, .(date), summarize, 
                       num_steps = sum(steps))

## Print an example of the summarized data.
        head(total_steps_per_day_df)
```

2. If you do not understand the difference between a histogram and a 
barplot, research the difference between them. Make a histogram of the
total number of steps taken each day.

```{r}
## Plot the histogram of total number of steps taken per day.
        hist(total_steps_per_day_df$num_steps, main = "HISTOGRAM OF TOTAL NUMBER OF STEPS PER DAY", 
        breaks = 20, xlab ="Number of steps taken", ylab= "Frequency")
```
        
3. Calculate and report the mean and median of the total number of steps 
taken per day

```{r}  
## Compute and print the mean and median of total number of steps taken per day.
        steps_per_day_summary1 <-c(mean(total_steps_per_day_df$num_steps), 
                                median(total_steps_per_day_df$num_steps))
        names(steps_per_day_summary1) <- c("Mean", "Median")
        head(steps_per_day_summary1)
```
   
## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis).

```{r}
# Compute the average steps per 5 minutes interval.
        avg_steps_per_interval_df <-  ddply(clean_data, .(interval), summarize, 
                       mean_steps = mean(steps))
## Plot the mean of steps take along the 5 minutes inteval time series.
        plot(avg_steps_per_interval_df, type = "l", xlab= "5 Minutes interval", 
             ylab= "Average steps per 5 minutes interval", col =2)
```

2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
```{r}
## Compute and print the 5 minutes interval with the maximum number of steps
## taken.
    max_steps_interval <- avg_steps_per_interval_df$interval[
                                which.max(avg_steps_per_interval_df$mean_steps)]
    

    print(paste0("5 minutes interval with more steps taken: ", max_steps_interval))
```
## Imputing missing values

Note that there are a number of days/intervals where there are missing values
(coded as NA). The presence of missing days may introduce bias into some calculations 
or summaries of the data.

1. Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs).

```{r}
## Compute and print the number of records with NA values.
        missing_values <- sum(!complete.cases(raw_data))
        print(paste0("Number of missing values: ", missing_values))
```

2. Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use the 
mean/median for that day, or the mean for that 5-minute interval, etc.

The strategy devised is to replace the NA values with randomly sampled values from
a copy of the dataset where the NAs have been removed. The seed used is equal to 
the total number of records in the dataset. set.seed(17568)

```{r}
## Randomly sample values from the clean dataset
        set.seed(17568)
        replacement_samples <- sample(x =clean_data$steps, 
        size = missing_values,
        replace = TRUE)
```

3. Create a new dataset that is equal to the original dataset but with the 
missing data filled in.

```{r}
## create a new dataset and replace the NA values with the randomly sampled ones.
        complete_data <- data.frame(raw_data)
        complete_data$steps[!complete.cases(raw_data)] <- replacement_samples
```

4. Make a histogram of the total number of steps taken each day and Calculate 
and report the mean and median total number of steps taken per day.

```{r}
## Summarize the number of steps taken per day.
        total_steps_per_day_df <- ddply(complete_data, .(date), summarize, 
                      num_steps = sum(steps))

## Plot histogram of number of steps taken each day employing the dataset with
## complete values.
        hist(total_steps_per_day_df$num_steps, main = "HISTOGRAM OF TOTAL NUMBER OF STEPS PER DAY", 
            breaks = 20, xlab= "Number of steps taken", ylab= "Frequency")
        
## Compute the mean and mediaan of steps taken per day using the dataset with 
## complete values.
        steps_per_day_summary1 <-c(mean(total_steps_per_day_df$num_steps), 
                                   median(total_steps_per_day_df$num_steps))
        names(steps_per_day_summary1) <- c("Mean","Median")
        
## Print the computed mean and median values.
        head(steps_per_day_summary1)
```
Do these values differ from the estimates from the first part of the assignment? 

    * Yes the mean and median differ from the ones computed ignoring the NAs.
  
What is the impact of imputing missing data on the estimates of the total daily number of steps?

    * The values of the mean and median for the filled out dataset are different than
    those computed when NAs are ignored. For the chosen seed, the mean and median
    result larger.
    
## Are there differences in activity patterns between weekdays and weekends?

    * Yes, there are differenes, during the weekends, the activity starts around 
    8:00am and spreads out across the whole day. However, 
    during weekdays, most of the walking activity happens prior the time to go to
    work, and then drops significantly during the office hours.
    
Create a new factor variable in the dataset with two levels – “weekday” 
and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
complete_data$date <- as.Date(complete_data$date ) 
complete_data_day_type <- cbind(complete_data, day_type = sapply(X = complete_data$date, 
                                                        FUN = function(x){
                                                            day = weekdays(x, abbreviate=TRUE)
                                                            day
                                                            if ((day== "Sat") ||
                                                                (day == "Sun"))
                                                                day = "Weekend"
                                                            else
                                                                day = "Weekday"
                                                        }, 
                                                        simplify = TRUE))
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute
interval (x-axis) and the average number of steps taken, averaged across all weekday
days or weekend days (y-axis). See the README file in the GitHub repository to see
an example of what this plot should look like using simulated data.

```{r}
library(plyr)
library(lattice)

steps_summary <- ddply(complete_data_day_type, .(day_type,interval), summarize, 
                       mean_steps = mean(steps))

xyplot(mean_steps~interval|day_type, data = steps_summary, type = "l", 
       layout= c(1,2), xlab ="Interval", ylab = "Number of steps")

```
